\documentclass[a4paper,11pt]{article}

\usepackage{fullpage}
%\usepackage[all]{xy}
\usepackage[utf8]{inputenc}
\usepackage{amsmath,amsthm}
\usepackage{amsfonts}
%\usepackage{algorithm}
\usepackage{graphicx}
%\usepackage[compatible,noend]{algpseudocode}
\usepackage{latexsym}
\usepackage{float}
\usepackage{color}
\usepackage{comment}

%%%%%%%%%%%%%%% COMMANDES %%%%%%%%%%%%%%%%%%%%%%%%%%%

\newcommand{\itemb}{\item[$\bullet$]}

%% Les ensembles de modelisation

\newcommand{\Fd}{\mathbb{F}}
\newcommand{\Znat}{\mathbb{Z}}
\newcommand{\Nnat}{\mathbb{N}}

%%%%%%%%%%%%%%% ENVIRONEMENTS %%%%%%%%%%%%%%%%%%%%%%%

%\newenvironment{correction}{\begin{comment}}{\end{comment}
%\newenvironment{correction}[1]{\begin{#1}}{\end{#1}}
%\includecomment{correction}

\theoremstyle{plain}
\newtheorem{lemme}{Lemma}
\newtheorem{algorithme}{ Algorithm }
\newtheorem{corolaire}{Corollary}
\newtheorem{propriete}{Property}
\newtheorem{proposition}{Proposition}
\newtheorem{theoreme} {Theorem}
\newtheorem{definition}{Definition}
\newtheorem{correction}{Answer}

\theoremstyle{definition}
\newtheorem{exercice}{Exercise}
%\theoremstyle{remark}
\newtheorem{remarque}{Remark}
\newtheorem{exemple}{Example}


%%%%%%%%%%%%%% DOCUMENTS %%%%%%%%%%%%%%%%%%%%%%%%%%%%


\begin{document}

\hbox{\raisebox{0.4em}{\vrule depth 0pt height 0.5pt width \textwidth}}
\noindent \textbf{University of Perpignan} \hfill  L3 Computer Science \\
 \hfill C. Legrand \hfill  Year \the\year \\
\smallskip
\begin{center}
{
  {\scshape \large Lab Work 1 - Security}  \\
 Malware
}



\end{center}
\hbox{\raisebox{0.4em}{\vrule depth 0pt height 0.9pt width \textwidth}}
%\hline

\bigskip

\bigskip


\begin{exercice}
\begin{enumerate}
\item To what extent are worms more dangerous than viruses?
\item Some worms that spread over the Internet do not cause any damage to infected machines. Why are they still harmful?
\item Name two techniques that allow viruses to evade detection by antivirus software.
\end{enumerate}
\end{exercice}

%\begin{comment}
\begin{correction}
\begin{enumerate}
\item The effectiveness of malware relies partly on its ability to spread quickly and easily. Since worms exploit networks, they can spread more easily and harm a larger number of machines.
\item Even if they do not cause any damage to machines, these worms still consume network resources for propagation at the expense of legitimate traffic, but more importantly, they exploit infected machines to carry out larger-scale attacks.
\item Examples include: polymorphism to avoid having an identifiable \textit{signature}, stealth techniques: the malicious code does not increase the size of the infected file, or it is hidden in a disk area marked as unavailable by the system, processes not appearing in the process list, etc.
\end{enumerate}
\end{correction}
%\end{comment}


\begin{exercice}
Consider the following program (in C language)
\begin{verbatim}
//... includes omitted
int main(int argc, char **argv)
{
  if(argc >= 1)
    {
       system("./cp argv[0] argv[1]");
    }
}
\end{verbatim}
Is this a virus?
\end{exercice}

%\begin{comment}
\begin{correction}
It is a virus in the sense that it duplicates its own code into a file (named \texttt{argv[1]}), and it is harmful in the sense that it overwrites that file.
\end{correction}
%\end{comment}


\begin{exercice}
In this exercise, we consider a variant of the W32/Beagle worm. This worm spreads as an email containing an attachment that is both compressed and encrypted. The password to decrypt the file is included in the message body. If the victim executes the file obtained after decompression using the provided password (which is an \texttt{.exe} file), the worm propagates by selecting its next victim from the current victim's address book.

Why is the compressed file encrypted if the password is provided in the message?

\end{exercice}

%\begin{comment}
\begin{correction}
The file is encrypted to evade detection by antivirus software that is unaware of this technique: the encrypted file changes with each encryption if the key changes, which provides polymorphism.
\end{correction}
%\end{comment}


%\begin{exercice}
%\'Ecrire un virus par écrasement de code (en Langage C, ou en un langage de script). Imaginer une/des stratégie pour que la taille du fichier infecté ne change pas.
%\end{exercice}


\begin{exercice}
Here is a description of the \texttt{Kork} malware
\begin{verbatim}
Kork
ALIAS:	Linux/Kork, Unix/Kork,

Kork  uses the known vulnerability in lpd service to propagate from a vulnerable
Linux system to another. This service is part of the default installation of Red
Hat Linux 7.0.

VARIANT:	Kork.A
If Kork.A finds a vulnerable host, it first creates two users, "kork" and "kork2",
to the system without a password. "kork2" user has a root priviledge. Kork also
adds an open shell to port 666.

Next it attempts to download a trojanized login and the main part of the trojan
from a web site. Since April 26th, 2001, neither of these files are available, so
it cannot replicate any further. However, already infected machines are able to
compromise other vulnerable machines by adding an open shell and users to the
system.

If the download is completed, Kork install an infected "/bin/login" and "/bin/ps"
to the system. It attemps to send sensitive system data propably to the author of
Kork. Original "/bin/ps" is copied to "/usr/bin/.ps" and the original "/bin/login"
is copied to "/bin/.login". Finally Kork starts to scan random Class-B subnets for
vulnerable hosts.
\end{verbatim}

\begin{enumerate}
\item What type of malware is Kork (logic bomb, Trojan horse, etc.)?
\item What are the potential targets of this malware? Briefly evaluate its impact.
\end{enumerate}
\end{exercice}

%\begin{comment}
\begin{correction}
\begin{enumerate}
\item We note that Kork propagates through the network and the \texttt{lpd} printing service on RedHat systems. It downloads Trojan horses (the \texttt{login} and \texttt{ps} utilities are corrupted). It also attempts to infect other machines. We observe that Kork replicates through the network, so it is partly a worm, but it also has Trojan horse functionality.
\item The potential targets are RedHat systems, which are rarely used by average users but are more commonly used for servers.
\end{enumerate}

\end{correction}
%\end{comment}


\begin{exercice}
Consider a report about a malicious executable named \texttt{Staog}.
\begin{verbatim}
NAME: Staog
SIZE: 4744

Staog spreads only under Linux operating system, infecting
Elf-style executables.Its first appearance was  in the fall of 1996,

Staog is written in assembler. It attempts to stay resident and
infect binaries as they are executed by any user. Stoag tries to
subvert root access via three known vulnerabilities (mount buffer
overflow, tip buffer overflow and one suidperl bug).

Staog contains several text strings, including:

Staog by Quantum / VLAD
/dev/kmemx/etc/mtab~
/sbin/mount
/tmp/t.dip
/bin/sh
/sbin/dip /tmp/t.dip
chatkey
/tmp/hs
#!/bin/sh\nchmod 666 /dev/kmem\n/tmp/hs
#!/usr/bin/suidperl -U\n$ENV{PATH}=\"/bin:/usr/bin\";
\n$>=0;$<=0;\nexec(\"chmod 666 /dev/kmem\");\n


Staog can be detected by searching all binaries for the following hex
search string:

215B31C966B9FF0131C0884309884314B00FCD80

Staog is not known to be in the wild at the time of this writing
(February 1997).
\end{verbatim}

\begin{enumerate}
\item What type of malware is Staog (logic bomb, Trojan horse, etc.)?
\item What are its potential targets and mode of operation? Briefly evaluate its impact.
\item What is its stealth strategy and its weakness?
\end{enumerate}

\end{exercice}

%\begin{comment}
\begin{correction}
\begin{enumerate}
\item Staog is a classic virus.
\item The targets are binary executables (ELF) on Linux systems. The impact is limited to Linux systems, which can cause significant damage to servers.
\item Its propagation method is relatively slow because it only spreads through file copying. It does not use the network.
\end{enumerate}
\end{correction}
%\end{comment}

\end{document}
